---
title: "Geospatial Analysis in R"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document explains the step-by-step geospatial analysis using R. The code below demonstrates the initialization of the R environment, data preprocessing, analysis, and visualization.

## Virtual Environment Initialization

We use the `renv` package to manage the virtual environment and `pacman` for streamlined package management.

```{r renv-setup}
library(renv)
library(pacman)

# Initialize renv virtual environment
if (!file.exists("renv.lock")) {
  renv::init() # Initialize the environment and create renv.lock if not already present
} else {
  renv::restore() # Restore the environment from renv.lock if it exists
}

# Install pacman if not already installed
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

# Use pacman to load and install libraries
pacman::p_load(
  here,
  sf,
  dplyr,
  ggplot2,
  scales,
  ggspatial,
  cowplot,
  tmap,
  tictoc,
  grid
)
```

## Data Loading

We load geospatial data and perform preprocessing for analysis. Shapefiles for Ecuador's political divisions and associated CSV data are read and prepared.

```{r data-loading}
# Working directory
here::here("geodata_tutorials", "01_sessions", "R-versions")

# Read shapefiles for Ecuador at different political levels
borders <- st_read(here("02_data", "shapefiles", "nivel-politico-4.shp"))
cantons <- st_read(here("02_data", "shapefiles", "nivel-politico-3.shp"))

# Drop and rename columns in the borders shapefile
borders <- borders %>%
  select(-TERRITORY_, -CATEGORY, -TYPE, -LEVEL_1) %>%
  rename(PARROQUIA = NAME, PROVINCIA = LEVEL_2, CANTON = LEVEL_3)

# Load additional data
df_final <- read.csv(here("02_data", "database", "df_final.csv"))
```

## Data Preparation

Data from different years (2017, 2022, and 2032) is extracted and merged for comparative analysis.

```{r data-preparation}
# Prepare data for 2017, 2022, and 2032
df_2017 <- df_final %>%
  filter(year == 2017) %>%
  select(parish_id, zonal_mean, PARROQUIA, PROVINCIA, CANTON) %>%
  rename(zonal_mean_2017 = zonal_mean)

df_2022 <- df_final %>%
  filter(year == 2022) %>%
  select(parish_id, zonal_mean, PARROQUIA, PROVINCIA, CANTON) %>%
  rename(zonal_mean_2022 = zonal_mean)

df_2032 <- df_final %>%
  filter(year == 2032) %>%
  select(parish_id, zonal_mean, PARROQUIA, PROVINCIA, CANTON) %>%
  rename(zonal_mean_2032 = zonal_mean)

# Merge data
df_merged <- df_2017 %>%
  left_join(df_2022, by = c("parish_id", "PARROQUIA", "PROVINCIA", "CANTON")) %>%
  left_join(df_2032, by = c("parish_id", "PARROQUIA", "PROVINCIA", "CANTON"))
```

## Computation of Avoided Conversion and Restoration

Avoided conversion and restoration metrics are calculated based on changes in zonal means over the years.

```{r metrics-computation}
# Compute avoided conversion
min_diff <- min(abs(df_merged$zonal_mean_2032 - df_merged$zonal_mean_2022), na.rm = TRUE)
max_diff <- max(abs(df_merged$zonal_mean_2032 - df_merged$zonal_mean_2022), na.rm = TRUE)

df_merged <- df_merged %>%
  mutate(
    avoided_conversion = case_when(
      is.na(zonal_mean_2032) | is.na(zonal_mean_2022) ~ NA,
      (zonal_mean_2032 - zonal_mean_2022) >= 0 ~ 0,
      max_diff > min_diff ~ (abs(zonal_mean_2032 - zonal_mean_2022) - min_diff) / (max_diff - min_diff),
      TRUE ~ 0
    )
  )

# Compute restoration
min_diff <- min(abs(df_merged$zonal_mean_2022 - df_merged$zonal_mean_2017), na.rm = TRUE)
max_diff <- max(abs(df_merged$zonal_mean_2022 - df_merged$zonal_mean_2017), na.rm = TRUE)

df_merged <- df_merged %>%
  mutate(
    restoration = case_when(
      is.na(zonal_mean_2022) | is.na(zonal_mean_2017) ~ NA,
      (zonal_mean_2022 - zonal_mean_2017) >= 0 ~ 0,
      max_diff > min_diff ~ (abs(zonal_mean_2022 - zonal_mean_2017) - min_diff) / (max_diff - min_diff),
      TRUE ~ 0
    )
  )
```

## Geospatial Processing

We merge data with shapefiles and ensure validity of geometries.

```{r geospatial-processing}
# Merge with shapefile
gdf_final <- df_merged %>%
  left_join(st_as_sf(borders), by = c("PARROQUIA", "PROVINCIA", "CANTON"))
gdf_final <- st_as_sf(gdf_final) # Convert back to an sf object

# Check for invalid geometries
st_is_valid(gdf_final)

# Fix invalid geometries
gdf_final <- st_simplify(gdf_final, dTolerance = 0.01)
gdf_final <- st_cast(gdf_final, "MULTIPOLYGON")
gdf_final <- st_make_valid(gdf_final)
```

## Visualization

A map visualizing the avoided conversion variable is created using `ggplot2`.

```{r visualization}
# Load boraders and water
ecuador_borders <- st_read(here("02_data", "shapefiles", "ecuador_borders.shp"))
ocean_borders <- st_read(here("02_data", "shapefiles", "ne_10m_ocean.shp")) 

# Set CRS if needed and reproject to raster CRS
# Replace 4326 with the appropriate EPSG code of your raster CRS
raster_crs <- st_crs(4326)

if (is.na(st_crs(ecuador_borders))) {
  ecuador_borders <- st_set_crs(ecuador_borders, 4326)
}
ecuador_borders <- st_transform(ecuador_borders, raster_crs)

if (is.na(st_crs(ocean_borders))) {
  ocean_borders <- st_set_crs(ocean_borders, 4326)
}
ocean_borders <- st_transform(ocean_borders, raster_crs)

# Replace NA values with 0 (or another appropriate value)
gdf_final$avoided_conversion[is.na(gdf_final$avoided_conversion)] <- 0
# Plot the avoided conversion variable
# Plot the avoided conversion variable
ggplot() +
  # Plot ocean borders first
  geom_sf(data = ocean_borders, fill = "#b2e8e8", alpha = 0.4, color = NA) +
  # Plot Ecuador borders
  geom_sf(data = ecuador_borders, fill = "#ffffff", color = "black", size = 0.5) +
  geom_sf(data = st_boundary(ecuador_borders), color = "black", size = 0.6) +
  # Plot avoided conversion with lighter, thinner borders
  geom_sf(data = gdf_final, aes(fill = avoided_conversion), color = "black", size = 0.3) +
  # Define the fill gradient with a horizontal color bar
  scale_fill_gradientn(
    colors = c("red", "white", "#00cc00"),
    limits = c(0, 1),
    name = "Avoided Conversion\n(0 = non-negative change)",
    guide = guide_colorbar(
      direction = "horizontal",   # Horizontal orientation
      barwidth = unit(5, "cm"),   # Smaller width
      barheight = unit(0.5, "cm"),# Smaller height
      title.position = "top",     # Title above
      title.hjust = 0.5           # Center-align title
    )
  ) +
  # Set coordinate limits to focus on Ecuador
  coord_sf(xlim = c(-81.5, -75), ylim = c(-5.5, 2.5)) +
  # Add country and ocean labels
  annotate("text", x = -76, y = -3.15, label = "PERU", size = 5, color = "black") +
  annotate("text", x = -76, y = 1, label = "COLOMBIA", size = 5, color = "black") +
  annotate("text", x = -80.9, y = 0, label = "PACIFIC\nOCEAN", size = 5, color = "#519ad0") +
  # Add a scale bar and north arrow
  annotation_scale(location = "bl", width_hint = 0.5) +
  annotation_north_arrow(location = "tl", which_north = "true",
                         style = north_arrow_fancy_orienteering()) +
  # Customize theme and legend position
  theme_minimal() +
  theme(
    legend.position = c(0.95, 0.1),    # Adjust legend position inside the plot
    legend.justification = c(1, 0),   # Bottom-right alignment
    legend.box.just = "right",        # Align legend elements to the right
    legend.title = element_text(size = 9, face = "bold"),  # Smaller title text
    legend.text = element_text(size = 7),                 # Smaller legend labels
    legend.background = element_rect(fill = "white", color = NA, size = 0.3),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1) # Add plot border
  ) +
  # Remove axis labels
  labs(x = NULL, y = NULL) +
  # Add plot title
  ggtitle("Avoided Conversion Variable (2022 to 2032)")